---
title: "Buckle-Up: Car Accident Data in Relation to Vehical Accidents/Deaths"
subtitle: "Project 2"
author: "Jefferson Ratliff"
title-slide-attributes:
  data-slide-number: none
format: revealjs
editor: visual
execute:
  echo: false
runtime: shiny
---

```{r, include=FALSE}
#| label: load-packages
#| include: false

library(tidyverse)
library(ggplot2)
library(plotly)
library(shiny)
library(AER)

#| label: setup
#| include: false

# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )

#| label: load-data
#| include: false
```

## Introduction

-   **Objective**: Analyzing the effects of mandatory seat belt laws on traffic fatalities and seat belt usage rates in the United States from 1983-1997.
-   **Enforcement Variables**:
    -   Primary Enforcement: Allows law enforcement officers to stop vehicles if a driver or passenger is not wearing a seat belt.
    -   Secondary Enforcement: Laws only allow police officers to issue tickets for seat belt violations if drivers have been pulled over for some other offense.

```{r}
library(AER)
data("USSeatBelts")
USSeatBelts <- na.omit(USSeatBelts)
```

------------------------------------------------------------------------

## Research Questions

1.  **Impact of Seat Belt Laws on Traffic Fatalities**:
    -   How do mandatory seat belt laws impact traffic fatalities across different US states over time?
2.  **Factors Influencing Seat Belt Usage Rates**:
    -   What factors influence seat belt usage rates in the United States, and how do these factors interact with one another?

------------------------------------------------------------------------

## Shiny Application 1: Seatbelt Usage by State Over Time

-   **All States**:
    -   Visualizes seatbelt usage across all states.
-   **State Breakdown**:
    -   Allows selection of a specific state to analyze seatbelt usage trends.

```{r}
ui <- navbarPage(
  title = "Seatbelt Usage by State",
  tabPanel(
    title = "All States",
    fluidRow(
      column(
        width = 12,
        plotlyOutput(outputId = "all_states_plot")
      )
    )
  ),
  tabPanel(
    title = "State Breakdown",
    fluidRow(
      column(
        width = 2,
        selectInput(inputId = "state_select", label = "Select State", 
                    choices = unique(USSeatBelts$state))
      ),
      column(
        width = 10,
        plotlyOutput(outputId = "plot")
      )
    )
  )
)


server <- function(input, output, session) {
  filtered_data <- reactive({
    filter(USSeatBelts, state == input$state_select)
  })
  output$plot <- renderPlotly({
    p <- ggplot(data = filtered_data(), aes(x = year, y = seatbelt, color = enforce, text = paste("State: ", input$state_select))) +
      geom_point() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
      labs(title = paste("Seatbelt Usage in", input$state_select))
    ggplotly(p) %>%
      layout(margin = list(l = 50, r = 50, b = 100, t = 100))
  })
  
  output$all_states_plot <- renderPlotly({
    p <- ggplot(data = USSeatBelts, aes(x = year, y = seatbelt, color = enforce, text = paste("State: ", state))) +
      geom_point() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
      labs(title = "Seatbelt Usage Across All States")
    ggplotly(p) %>%
      layout(margin = list(l = 50, r = 50, b = 100, t = 100))
  })
}

shinyApp(ui = ui, server = server)
```

------------------------------------------------------------------------

## Shiny Application 2: Impact of Seat Belt Laws on Traffic Fatalities

-   **All States**:
    -   Shows the mean traffic fatalities across all states over time.
-   **State Breakdown**:
    -   Allows selection of a specific state to analyze traffic fatalities over time.

```{r}
USSeatBelts$year <- as.numeric(as.character(USSeatBelts$year))

ui <- fluidPage(
  titlePanel("Impact of Seat Belt Laws on Traffic Fatalities"),
  tabsetPanel(
    tabPanel("All States",
             sidebarLayout(
               sidebarPanel(
                 sliderInput("years_all", "Select Years:",
                             min = min(USSeatBelts$year),
                             max = max(USSeatBelts$year),
                             value = c(min(USSeatBelts$year), max(USSeatBelts$year)))
               ),
               mainPanel(
                 plotOutput("fatalitiesPlot_all")
               )
             )
    ),
    tabPanel("State Breakdown",
             sidebarLayout(
               sidebarPanel(
                 selectInput("state", "Select State:",
                             choices = unique(USSeatBelts$state),
                             selected = unique(USSeatBelts$state)[1]),
                 sliderInput("years", "Select Years:",
                             min = min(USSeatBelts$year),
                             max = max(USSeatBelts$year),
                             value = c(min(USSeatBelts$year), max(USSeatBelts$year)))
               ),
               mainPanel(
                 plotOutput("fatalitiesPlot")
               )
             )
    )
  )
)

server <- function(input, output) {
  
  output$fatalitiesPlot <- renderPlot({
    selected_data <- USSeatBelts[USSeatBelts$state == input$state & 
                                   USSeatBelts$year >= input$years[1] & 
                                   USSeatBelts$year <= input$years[2], ]
    plot(selected_data$year, selected_data$fatalities, 
         xlab = "Year", ylab = "Fatalities",
         main = paste("Traffic Fatalities in", input$state, "Over Time"))
    lines(selected_data$year, selected_data$fatalities, col = "blue")
  })
  
  output$fatalitiesPlot_all <- renderPlot({
    mean_data <- aggregate(fatalities ~ year, data = USSeatBelts, FUN = mean)
    plot(mean_data$year, mean_data$fatalities, 
         xlab = "Year", ylab = "Mean Fatalities",
         main = "Mean Traffic Fatalities Across All States Over Time")
    lines(mean_data$year, mean_data$fatalities, col = "blue")
  })
}

shinyApp(ui = ui, server = server)
```

------------------------------------------------------------------------

## Data Analysis

-   **Key Findings**:
    -   Strong evidence suggests that the implication and enforcement of seat belt laws have effectively increased both fatalities and seat belt usage.
    -   Rapid growth in seat belt use and enforcement observed over a short period.

------------------------------------------------------------------------

## Thank You!

-   **Questions?**
